CytoGAN: Generative Modeling of Cell Images

### 2017-11-28 paper/paper.tex
Updates to paper and poster

\author{
  Peter Goldsborough\thanks{This work has been conducted as part of an internship at the Imaging Platform of the Broad Institute.}\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{pgoldsbo@broadinstitute.org} \\
  \And
  Nick Pawlowski \\
  Department of Computing \\
  Imperial College London \\
  London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C Caicedo \& Shantanu Singh \& Anne E Carpenter\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{\{jccaicedo, shsingh, anne\}@broadinstitute.org} \\
}

@@

    \foreach \x/\l in {0/Convolution,%
                       1/Convolution,%
                       2/Convolution,%
                       3/{\hspace{1mm}Dense$(\ell)$\textsuperscript{\textdagger}},%
                       4/{Dense$(1)$}%
                       } {
      \path ({5.8 + \x * 0.65}, 0)
            coordinate [box, text width=0.3cm, text height=2cm] (l\x)
            node [rotate=90] {\l};
    }

    % Footnote
    \draw (7, -1.5)
          node {\footnotesize{\textdagger} Representations};

@@

Figure \ref{fig:generated} shows examples of images generated with LSGAN, WGAN
and DCGAN architectures alongside real images. The synthetic images,
particularly those produced by LSGAN, are not only highly detailed and
realistic, but also consistent with their biological nature. For example, it is
characteristic that $\beta$-Tubulin forms a circular halo cradling the nucleus.
This property is maintained clearly in all generated images.

@@

This work covers only a small fraction of possible applications of GANs to the domain of microscopy images. For example, we hope that future work investigates BiGANs\cite{donahue2016adversarial} or other novel solutions to infer latent features with GANs. We believe that improved inference, combined with the interpretable nature of the GAN framework, may enable simulated experiments by performing algebra with vectors corresponding to cell lines, diseases or other perturbations. Finally, we note that deep learning generally flourishes with larger amounts of data. Thus, we conjecture that the performance of the outlined approaches will improve with larger datasets.

>>>>

\author{
  Peter Goldsborough\thanks{This work has been conducted as part of an internship at the Imaging Platform of the Broad Institute.}\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{pgoldsbo@broadinstitute.org} \\
  \And
  Nick Pawlowski \\
  Department of Computing \\
  Imperial College London \\
  London, UK \\
  \texttt{n.pawlowski16@imperial.ac.uk} \\
  \AND
  Juan C Caicedo \& Shantanu Singh \& Anne E Carpenter\\
  Imaging Platform \\
  Broad Institute of MIT and Harvard \\
  Cambridge, MA, USA \\
  \texttt{\{jcaicedo, shsingh, anne\}@broadinstitute.org} \\
}

@@

    \foreach \x/\l in {0/Convolution,%
                       1/Convolution,%
                       2/Convolution,%
                       3/{\hspace{1mm}Dense$(\ell)$\textsuperscript{$\star$}},%
                       4/{Dense$(1)$}%
                       } {
      \path ({5.8 + \x * 0.65}, 0)
            coordinate [box, text width=0.3cm, text height=2cm] (l\x)
            node [rotate=90] {\l};
    }

    % Footnote
    \draw (7, -1.5)
          node {\footnotesize{\textsuperscript{$\star$}}\hspace{-1pt} Representations};

@@

Figure \ref{fig:generated} shows examples of images generated with LSGAN, WGAN
and DCGAN architectures alongside real images. The synthetic images,
particularly those produced by LSGAN, are not only highly detailed and
realistic, but also consistent with their biological nature. For example, it is
characteristic that $\beta$-Tubulin forms a circular halo cradling the nucleus.
This property is maintained clearly in most generated images.

@@

This work covers only a small fraction of possible applications of GANs to the domain of microscopy images. For example, we hope that future work investigates BiGANs\cite{donahue2016adversarial} or other novel solutions to infer latent features with GANs. We believe that improved inference, combined with the interpretable nature of the GAN framework, may enable simulated experiments by performing algebra with vectors corresponding to cell lines, diseases or other perturbations.

### 2017-11-28 paper/poster/poster.tex
Updates to paper and poster

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{relsize}

@@

\usepackage[
  backend=biber,
  sorting=nty,
  style=ieee,
  minbibnames=1,
  maxbibnames=1,
  maxcitenames=1,
  mincitenames=1,
  defernumbers=true]{biblatex}
\addbibresource{../lit.bib}

@@

\background{
  \begin{tikzpicture}[remember picture,overlay]%
    \draw (12.75, 7.5) node {\includegraphics[scale=1]{tiled}};
    \draw (12.75, 24) node {\includegraphics[scale=1]{tiled}};

@@

    \draw[white, fill=white, rounded corners=1pt]
         (current page.north east)+(-3.2, -0.5) rectangle ++(-0.3, -3.2);

@@

  \vspace{2mm}
  {
    \smaller[1]
    \textsuperscript{$\star$} Broad Institute of MIT and Harvard \hspace{3mm}%
    \textdagger\hspace{2pt} Technical University Munich \hspace{3mm}%
    $\ddagger$ Imperial College London
  }
}
% Logos
{
  \begin{tabular}{l}
    \includegraphics[width=0.1\textwidth]{broad-logo}\\
    \vspace{-2mm}\\
    \includegraphics[width=0.1\textwidth]{imperial-logo}
    \vspace{-2mm}\\
    \includegraphics[width=0.1\textwidth]{tum-logo}\\
  \end{tabular}
}

@@

Current approaches are divided between a) classical image processing, requiring manual fine-tuning and expert knowledge b) transfer learning with deep CNNs trained on miscellaneous objects, allowing no domain-specific adaptation.

@@

Instead, we model cells with \emph{Generative Adversarial Networks} (GANs) for both representation learning and cell synthesis. Advantages of our approach are:
\vspace{2mm}
\begin{itemize}
\compresslist
\item \textbf{Unsupervised} representation learning, which can readily incorporate new data;
\item \textbf{Specialization} to training data, allowing identification of the semantic relations between biologically meaningful channels;
\item \textbf{Interpretability} of learned representations and useful visualization that help translate data variations into meaningful biological phenotypes.
\end{itemize}
}

@@

Shown above are cells synthesized with LSGAN, WGAN, BEGAN and BiGAN
architectures alongside a real cell. The synthesized images are not only highly
detailed and realistic, but also consistent with their biological nature. For
example, it is characteristic that $\beta$-Tubulin (green channel) forms a
circular halo cradling the nucleus. This property is maintained clearly in all
generated images. }

@@

GANs are known to learn low manifolds of their input priors, such that
interpolation between two points $\mathbf{z}_1, \mathbf{z}_2$ drawn from the noise space $P_{\text{noise}}$ results in visually smooth
transitions in generated images. We confirm that this property holds for
microscopy cell images.

@@

Bidirectional GANs (BiGAN\cite{}) extend the vanilla DCGAN architecture with an
explicit encoder network that enables both synthesis of images from noise
\emph{and} inference of latent representations, in such a way that the noise and
latent space are fused. This means we can encode real cell images, perturb,
transform or, in this case, interpolate between them, and re-synthesize the
resulting vectors.

@@

When the noise and learned latent space are fused (BiGAN), algebra on real cell images becomes possible. This open the door to discovery of a whole new class of highly interpretable and biologically valuable relationships, such as:

\begin{tikzpicture}
  \draw (-8, 0);
  \draw (0, 0) node [text width=10cm] {
  \begin{itemize}
    \compresslist
    \item $\text{emetine}_{1.0} - \text{emetine}_{0.3} + \text{taxol}_{0.3} \stackrel{?}{=} \text{taxol}_{1.0}$
    \item $\text{Protein Synthesis} - \text{Protein Degradation} \stackrel{?}{=} \text{DMSO}$
    \item $\text{Kinase Inhibitor} - \text{DMSO} + \text{DMSO}' \stackrel{?}{=} \text{Kinase Inhibitor}$
  \end{itemize}
  };
\end{tikzpicture}
\vspace{2.5pt}
}

\headerbox{Representation Learning for Morphological Profiling}{name=rep, span=2, column=1, row=1, below=algebra}{

We investigate the ability of GANs to learn vector representations of cell
images and evaluate their quality quantitatively at the task of
mechanism-of-action (MOA) prediction. We train a variety of GAN models on 1.3
million microscopy images of cells perturbed with particular drug compounds,
average signatures across treatments and predict the MOA via nearest-neighbor
classification.

\vspace{2mm}
\begin{tikzpicture}
  \draw (-2, 0);
  \draw (0, 0.3) node {\includegraphics[scale=0.2, trim={4cm 3.5cm 8cm 3cm}, clip]{latent-space}};
  \draw (7.75, 1.2) node [text width=11cm, align=justify] {
  For WGAN and LSGAN we interpret the activations of the penultimate dense
  layer as a source of representations. BEGAN an BiGAN architectures already have explicit inference modules. We find that while GANs
  are superior to autoencoders, they are not yet competitive with existing
  classical or deep transfer-learning techniques.
  };

@@

\headerbox{Code \& Data}{name=code, column=0, row=1, below=intro}{Code and experiments are available at \url{github.com/goldsborough/cytogan}. Our data is the BBBC021 dataset from the Broad BioImage Benchmark Collection (\url{data.broadinstitute.org/bbbc/BBBC021/}).}

\headerbox{Acknowledgements}{name=ack, column=0, row=2, below=code}{
We thank Mike Ando and Google, Inc. for providing computational resources to accelerate our research. We also thank all members of the Broad Imaging Platform for fruitful discussion and their continuous guidance and support.
}
\headerbox{References}{name=ref, column=0, row=3, below=ack}{
\printbibliography
}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[american]{babel}

\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{relsize}

@@

\usepackage[
  backend=biber,
  sorting=nty,
  style=ieee,
  minbibnames=1,
  maxbibnames=1,
  maxcitenames=1,
  mincitenames=1,
  defernumbers=true,
  natbib=true]{biblatex}
\addbibresource{../lit.bib}

@@

\background{
  \begin{tikzpicture}[remember picture,overlay]%
    % \draw (12.75, 7.5) node {\includegraphics[scale=1]{tiled}};
    % \draw (12.75, 24) node {\includegraphics[scale=1]{tiled}};

@@

    \draw[white, fill=white, rounded corners=1pt]
         (current page.north east)+(-3.3, -0.5) rectangle ++(-0.4, -3.2);

@@

  \vspace{2mm}
  {
    \smaller[1]
    \aff Broad Institute of MIT and Harvard \hspace{3mm}%
    \textdagger\hspace{2pt} Technical University Munich \hspace{3mm}%
    $\ddagger$ Imperial College London
  }
}
% Logos
{
  \begin{tabular}{l}
    \includegraphics[width=0.11\textwidth]{broad-logo}\\
    \vspace{-2mm}\\
    \includegraphics[width=0.11\textwidth]{imperial-logo}
    \vspace{-2mm}\\
    \includegraphics[width=0.11\textwidth]{tum-logo}\\
  \end{tabular}
}

@@

Current approaches are divided between \textbf{a}) classical image processing, requiring manual fine-tuning and expert knowledge, and \textbf{b}) transfer learning with deep CNNs trained on miscellaneous objects, allowing little to no domain-specific adaptation.

@@

Instead, we model cells with \emph{Generative Adversarial Networks} (GANs) for both representation learning and cell image synthesis. Advantages of our approach include:
\vspace{2mm}
\begin{itemize}
\compresslist
\item \textbf{Unsupervised} representation learning, which can readily incorporate new data;
\item \textbf{Specialization} to training data, allowing identification of the semantic relations between biologically meaningful channels;
\item \textbf{Interpretability} of learned representations and useful visualization that help translate data variations into meaningful biological phenotypes.
\end{itemize}
}

@@

Shown above are cells synthesized with LSGAN, WGAN, BEGAN and BiGAN
architectures alongside a real cell. The synthesized images are not only highly
detailed and realistic, but also consistent with their biological nature. For
example, it is characteristic that $\beta$-Tubulin (green channel) forms a
circular halo cradling the nucleus. This property is maintained clearly in most
generated images. }

@@

GANs are known to learn low manifolds of their input priors, such that
interpolation between two points $\mathbf{z}_1, \mathbf{z}_2$ drawn from the noise space $P_{\text{noise}}$ results in visually smooth
transitions in generated images \cite{radford2015unsupervised}. We confirm that this property holds for microscopy cell images.

@@

Bidirectional GANs (BiGANs) \cite{donahue2016adversarial} extend the vanilla DCGAN \cite{radford2015unsupervised} architecture with an
explicit encoder network that enables both synthesis of images from noise
\emph{and} inference of latent representations, in such a way that the noise and
latent space are fused. This means we can encode real cell images, perturb,
transform or in this case interpolate between them and re-synthesize the
resulting images.

@@

When the noise and learned latent space are fused (BiGAN), algebra on real cell
images becomes possible. This opens the door to discovery of a whole new class
of highly interpretable and biologically valuable relationships, such as:

\vspace{1mm}
\begin{tikzpicture}
  \draw (-8, 0);
  \draw (0, 0) node [text width=10cm] {
  \begin{itemize}
    \compresslist
    \item $\text{emetine}_{1.0} - \text{emetine}_{0.3} + \text{taxol}_{0.3} \stackrel{?}{=} \text{taxol}_{1.0}$
    \item $\text{Protein Synthesis} - \text{Protein Degradation} \stackrel{?}{=} \text{DMSO}$
    \item $\text{Kinase Inhibitor} - \text{DMSO} + \text{DMSO}' \stackrel{?}{=} \text{Kinase Inhibitor}$
  \end{itemize}
  };
\end{tikzpicture}
\vspace{-2mm}
}
\headerbox{Representation Learning for Morphological Profiling}{name=rep, span=2, column=1, row=1, below=algebra}{

We investigate the ability of GANs to learn high quality vector representations
of cell images and evaluate them quantitatively at the task of
mechanism-of-action (MOA) prediction. We train a variety of GAN models on 1.3
million microscopy images of cells perturbed with particular drug compounds,
average signatures across treatments and predict the MOA via nearest-neighbor
classification.

\vspace{2mm}
\begin{tikzpicture}
  \draw (-2, 0);
  \draw (0, 0.3) node {\includegraphics[scale=0.2, trim={4cm 3.5cm 8cm 3cm}, clip]{latent-space}};
  % \draw (7.75, 1.2) node [text width=11cm, align=justify] {
  % For LSGAN we interpret the activations of the penultimate dense
  % layer as a source of representations. BiGAN has an explicit inference module and requires no further modification. We find that while GANs
  % are superior to autoencoders, they are not yet competitive with existing
  % classical or deep transfer-learning techniques at this task.
  % };

  \draw (7.75, 1.2) node {\scalebox{0.8}{
    \begin{tikzpicture}[thick]
      \tikzset{box/.style={
        draw, rectangle, rounded corners=0.8pt,
        text width=2.2cm, text height=0.75cm}}
      \tikzset{node/.style={draw, circle, inner sep=0.35cm}}

      % Spacer
      \draw (10.5, 0);

      % Model
      \path (0, 0) coordinate [draw, circle, inner sep=5pt]
            (z) node {$\mathbf{z}$};

      \path (2, 0) coordinate [box] (G) node {Generator};

      \path (4.3, -0.8) coordinate [node] (f) node {$G(\mathbf{z})$};
      \path (4.3, +0.8) coordinate [node] (r) node {$\mathbf{x}$};

      \path (9.5, 0) coordinate [draw, circle, inner sep=7pt]
            (D) node {$\hspace{1pt}\mathbf{D}$};

      % Anchor for real and fake arrows.
      \path (5, 0) coordinate (_);

      \foreach \x/\l in {0/Convolution,%
                         1/Convolution,%
                         2/Convolution,%
                         3/{\hspace{1mm}Dense$(\ell)$\textsuperscript{$\star$}},%
                         4/{Dense$(1)$}%
                         } {
        \path ({5.8 + \x * 0.65}, 0)
              coordinate [box, text width=0.5cm, text height=2cm] (l\x)
              node [rotate=90] {\l};
      }

      % Footnote
      \draw (7, -1.5)
            node {\footnotesize{\textsuperscript{$\star$}}\hspace{-1pt} Representations};

      % Edges
      \draw (f) -- (_);
      \draw (r) -- (_);
      \draw [->] (_) -- (l0);
      \draw [->] (l4) -- (D);
      \draw [->] (z) -- (G);
      \draw [->] (G) edge[out=0] (f);
    \end{tikzpicture}
  }};

@@

\headerbox{Code \& Data}{name=code, column=0, row=1, below=intro}{
Code and experiments are available at \url{github.com/goldsborough/cytogan}. Our data is the BBBC021 dataset from the Broad BioImage Benchmark Collection (\url{data.broadinstitute.org/bbbc/BBBC021}).

% REPLACE WITH PAPER!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
\vspace{-3mm}
\begin{center}
  \includegraphics[scale=0.174]{qr-code}\hspace{1cm}\includegraphics[scale=0.174]{qr-code}

  \vspace{-3mm}
  \hspace{0.1cm}\texttt{Code}\hspace{2.6cm}\texttt{Paper}
\end{center}

% \includegraphics[scale=0.1]{qr-paper}
}

\headerbox{Acknowledgements}{name=ack, column=0, row=2, below=code}{
We thank Mike Ando and Google, Inc. for providing computational resources to accelerate our research. We also thank all members of the Broad Imaging Platform for fruitful discussion and their continuous guidance and support.
}

\headerbox{References}{name=ref, column=0, row=3, below=ack}{
\vspace{-5mm}
\printbibliography
}
